Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-09-07 paper/paper.tex
Continue

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

@@

\newcommand{\fmAnd}[2]{#1 $\boldsymbol{\land}$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNeg}[1]{$\boldsymbol{\neg}$ #1}
\newcommand{\fmAtom}{\textbf{A}}

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol} provides some further context to this entry.
The entry declares a type \lstinline!V! that is equipped with a function \lstinline!elts :: V $\Rightarrow$ V set! that maps a set to its elements.
\todo[inline]{Use itemize for different functions}
Even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
Only if the class is small enough, as indicated by the polymorphic predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, it is indeed a set.
Inversely, this predicate characterises those sets that can be turned into a set of type \lstinline!V! using the polymorphic function \lstinline!vset :: 'a set $\Rightarrow$ V!.
The type \lstinline!V! comes with all the usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$).
The empty set coincides with the ordinal $0$ so it is denoted by \lstinline!0 :: V!.

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \textit{closed} if all of its branches are closed.
\todo[inline]{Move closed}

@@

definition subterms' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms' b $\equiv$
    subterms (last b) $\cup$ 
    Var ` (wits b - wits' b)
\end{lstlisting}
We aim to construct a syntactic model that is derived from the membership literals \lstinline!(s $\inS$ t) $\in$ set b!.
To this end, we construct a graph whose vertices are set terms and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} where a graph is represented as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source respectively target vertex.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` wits' b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs,
     &$\phantom{\llparenthesis}$& arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}
The model is defined in terms of a \textit{realization} function relative to this graph.
Observe that the vertices are partitioned into two sets, a collection of witnesses \lstinline!Var ` wits' b! and the rest of the set terms \lstinline!subterms' b!, which are treated differently by the realization.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realization of a vertex is defined as the union of the realizations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` wits' b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!vset {I x}!, which, together with the cardinality constraint, ensures that realizations are distinct between those witnesses and the set terms.  
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

@@

\subsection{Realisation of an Open Branch}
Remember that for completeness, we need to show that the realization function for an open and saturated branch \lstinline!b! actually constitutes a model for all formualae in the branch.
We start by verifying that the realization function models all literals in the branch; more formally, the following propositions hold: 
\begin{enumerate}[label=(\roman*)]
  \item\label{it:at_mem} We have \lstinline!realize s $\in$ elts (realize t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realize s $=$ realize t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realize s $\neq$ realize t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realize s $\notin$ elts (realize t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \lstinline!lemma_2!, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ wits' b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \lstinline!lemma_2!.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\inS$ subterms' b! must hold.
  Now, assume for contradiction that \lstinline!realize s $\neq$ realize t! which implies that \lstinline!elts (realize s) $\neq$ elts (realize t)!.
  Without loss of generality --- the other case is symmetric ---, we obtain an \lstinline!a! such that
\begin{lstlisting}
  a $\in$ elts (realize s) $\land$ a $\notin$ elts (realize t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realize!, we obtain an \lstinline!u! such that \lstinline!a = realize u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ b! and the openness of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realize u $\in$ elts (realize t)
        = a $\in$ elts (realize t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!a $\notin$ elts (realize t)!.
\end{proof}

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in Section~\ref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see Section~\ref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by rule induction suffices to obtain the lemma below.
\begin{lstlisting}
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b'
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values, hence modifying the model, as we will see in the proof below.
\begin{lstlisting}[belowskip=0pt]
lemma fexpands_sound:
  assumes &\fexpands{bs'}{b}&
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'.
          $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We perform a case analysis on the derivation \lstinline!&\fexpands{bs'}{b}&!.
  As all cases but that of the last rule are straightforward, we only consider this last case where we have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t],
         [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!, i.e.\ \lstinline!x $\notin$ vars b! and in particular \lstinline!x $\notin$ vars s $\cup$ vars t!.
  Another consequence of applying the last rule is that \lstinline!s $\neqS$ t! is in \lstinline!b!, which, because \lstinline!M! is a model, implies that \lstinline!&\Ist& M s $\neq$ &\Ist& M t!.
  This inequality manifests itself through some \lstinline!y! with
\begin{lstlisting}
  &\phantom{$\lor$}& y $\in$ elts (&\Ist& M s) $\land$ y $\notin$ elts (&\Ist& M t)
  $\lor$ y $\in$ elts (&\Ist& M t) $\land$ y $\notin$ elts (&\Ist& M s).
\end{lstlisting}
  We now update the assignment \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that values of both \lstinline!s! and \lstinline!t! are unchanged under this update as \lstinline!x $\notin$ vars s! and \lstinline!x $\notin$ vars t!.
  Considering our assumption about \lstinline!y! and the fact that \lstinline!&\Ist& M' (Var x) = y!, we infer that
\begin{lstlisting}
  $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set b'. M' $\models$ $\psi$.
\end{lstlisting}
  In other words, due to the disjunction, it holds that \lstinline!M'! is a model for one \lstinline!b' $\in$ bs!.
  Thus, we are left with proving that \lstinline!M'! is a model for \lstinline!b!.
  This is indeed the case since the assumption \lstinline!M $\models$ $\psi$! is equivalent to \lstinline!M' $\models$ $\psi$! for all \lstinline!$\psi$ $\in$ set b! given that \lstinline!x! is fresh with respect to b.
\end{proof}

\section{Upper Bounding the Cardinality of a Branch}
Soundness and completeness of the calculus are the building blocks with which we establish partial correctness of Algorithm~\ref{fig:alg}.
Consequently, we still need to give a termination proof of Algorithm~\ref{fig:alg} to obtain total correctness of the algorithm.
The core of our argument is that the number of distinct formulae in a branch that can be derived with the expansion rules is finite.
In particular, we claim that the number of distinct formulae has the following upper bound.
\begin{lstlisting}
  &\phantom{$\leq$}& card (set b)
  &\textover{$=$}{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$} +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& 16 * (card (subterms (last b)))$^4$
\end{lstlisting}
We justify both inequalities separately, starting with the first one whose justification hinges on an invariant:
Its justification hinges on the invariant that the only new subformulae introduced by the expansion rules introduce

@@

>>>>

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

% Autoref names
\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Section}

@@

\newcommand{\fmAnd}[2]{#1 $\boldsymbol{\land}$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNegSymbol}{\boldsymbol{\neg}}
\newcommand{\fmNeg}[1]{$\fmNegSymbol$ #1}
\newcommand{\fmAtom}{\textbf{A}}

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol} provides some further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
      This class only constitutes a set if the class is small enough.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a set is small.
Inversely, this predicate characterises those sets that can be turned into a set of type \lstinline!V! using the polymorphic function \lstinline!vset :: 'a set $\Rightarrow$ V!.
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$ so it is denoted by \lstinline!0 :: V!.
\end{itemize}

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \textit{closed} if all of its branches are closed.

@@

definition subterms' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms' b $\equiv$
    subterms (last b) $\cup$ 
    Var ` (wits b - wits' b)
\end{lstlisting}
We aim to construct a syntactic model that is derived from the membership literals \lstinline!(s $\inS$ t) $\in$ set b!.
To this end, we construct a graph whose vertices are set terms and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} where a graph is represented as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source respectively target vertex.
\todo[inline]{Explain arrow notation}
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` wits' b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs,
     &$\phantom{\llparenthesis}$& arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}
The model is defined in terms of a \textit{realization} function relative to this graph.
Observe that the vertices are partitioned into two sets, a collection of witnesses \lstinline!Var ` wits' b! and the rest of the set terms \lstinline!subterms' b!, which are treated differently by the realization.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realization of a vertex is defined as the union of the realizations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` wits' b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!vset {I x}!, which, together with the cardinality constraint, ensures that realizations are distinct between those witnesses and the set terms.  
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

@@

\subsection{Realisation of an Open Branch}
Remember that for completeness, we need to show that the realization function for an open and saturated branch \lstinline!b! actually constitutes a model for all formualae in the branch.
We start by verifying that the realization function models all literals in the branch; more formally, the following propositions hold: 
\begin{enumerate}[label=(\roman*)]
  \item\label{it:at_mem} We have \lstinline!realize s $\in$ elts (realize t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realize s $=$ realize t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realize s $\neq$ realize t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realize s $\notin$ elts (realize t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \lstinline!lemma_2!, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ wits' b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \lstinline!lemma_2!.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\inS$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  &\phantom{$\Longrightarrow$}& realize s $\neq$ realize t
  $\Longrightarrow$ elts (realize s) $\neq$ elts (realize t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric ---, we obtain an \lstinline!a! such that
\begin{lstlisting}
  a $\in$ elts (realize s) $\land$ a $\notin$ elts (realize t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realize!, we obtain an \lstinline!u! such that \lstinline!a = realize u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ b! and the openness of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realize u $\in$ elts (realize t)
        = a $\in$ elts (realize t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!a $\notin$ elts (realize t)!.
\end{proof}

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in \autoref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see \autoref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by rule induction suffices to obtain the lemma below.
\begin{lstlisting}
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b'
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values; hence, we modify the model as we will see in the proof below.
\begin{lstlisting}[belowskip=0pt]
lemma fexpands_sound:
  assumes &\fexpands{bs'}{b}&
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'.
          $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We perform a case analysis on the derivation \lstinline!&\fexpands{bs'}{b}&!.
  As all cases but that of the last rule are straightforward, we only consider this last case where we have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t],
         [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!, i.e.\ \lstinline!x $\notin$ vars b! and in particular \lstinline!x $\notin$ vars s $\cup$ vars t!.
  Another consequence of applying the last rule is that \lstinline!s $\neqS$ t! is in \lstinline!b!, which, because \lstinline!M! is a model, implies that \lstinline!&\Ist& M s $\neq$ &\Ist& M t!.
  This inequality manifests itself through some \lstinline!y! with
\begin{lstlisting}
  &\phantom{$\lor$}& y $\in$ elts (&\Ist& M s) $\land$ y $\notin$ elts (&\Ist& M t)
  $\lor$ y $\in$ elts (&\Ist& M t) $\land$ y $\notin$ elts (&\Ist& M s).
\end{lstlisting}
  We now update the assignment \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that values of both \lstinline!s! and \lstinline!t! are unchanged under this update as \lstinline!x $\notin$ vars s! and \lstinline!x $\notin$ vars t!.
  Considering our assumption about \lstinline!y! and the fact that \lstinline!&\Ist& M' (Var x) = y!, we infer that
\begin{lstlisting}
  $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set b'. M' $\models$ $\psi$.
\end{lstlisting}
  In other words, due to the disjunction, it holds that \lstinline!M'! is a model for one \lstinline!b' $\in$ bs!.
  Thus, we are left with proving that \lstinline!M'! is a model for \lstinline!b!.
  This is indeed the case since the assumption \lstinline!M $\models$ $\psi$! is equivalent to \lstinline!M' $\models$ $\psi$! for all \lstinline!$\psi$ $\in$ set b! given that \lstinline!x! is fresh with respect to b.
\end{proof}

\section{Upper Bounding the Cardinality of a Branch}
Soundness and completeness of the calculus are the building blocks with which we establish partial correctness of \autoref{fig:alg}.
Consequently, we still need to give a termination proof to ascertain total correctness.
The core of our argument is that the number of distinct formulae that can be derived with the expansion rules is finite.
In particular, we claim that the following upper bound holds.
\begin{lstlisting}
  &\phantom{$\leq$}& card (set b)
  &\textover{$=$}{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$} +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& 16 * (card (subterms (last b)))$^4$
\end{lstlisting}
We treat both inequalities separately, starting with the first one whose justification hinges on an invariant: 
every new composite subformula that is introduced by an expansion rule is the negation of a subformula that already occurs in \lstinline!last b!.
Written as a formula, we have
\begin{lstlisting}
  &\phantom{$\subseteq$}& {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\subseteq$ subfms (last b) $\cup$ $\fmNegSymbol$ ` subfms (last b).
\end{lstlisting}
Confirming that this is an invariant is an easy exercise of going through each of the propositional rules in Table TODO and the first two branching rules in Table TODO.
Thus, we have
\begin{lstlisting}
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\leq$ card (subfms (last b)) +
  &\phantom{$\leq$}& card ($\fmNegSymbol$ ` subfms (last b))
  $\leq$ 2 * card (subfms (last b))
\end{lstlisting}
which is the first inequality we asserted.

Concerning the second inequality, first observe that each literal contains an atom, implying that the set of literals is at most twice as large as the set of atoms in a branch. 
In addition, an atom is formed by combining two subterms occuring in the branch with either \lstinline!$\inS$! or \lstinline!$\eqS$!, ultimately yielding
\begin{lstlisting}
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card ($\bigcup$(atoms ` set b))
  $\leq$ 2 * 2 * (card (subterms b))$^\text{2}$.
\end{lstlisting}
Since no expansion rule introduces new subterms, we know the cardinality of the subterms in a branch \lstinline!b! is limited by the subterms of \lstinline!last b! plus the number of witnesses added to the branch.
Regarding an upper bound for the witnesses, recognise that the last branching rule requires \lstinline!s $\neqS$ t! to be in \lstinline!b! for some \lstinline!s,t $\in$ subterms (last b)!.
The subsumption condition then prohibits that the rule can be applied again for the combination of \lstinline!s! and \lstinline!t!, thus yielding an upper bound of \lstinline!(card (subterms (last b)))$^\text{2}$!.
This allows us to finish the calculation that justifies the second inequality:
\begin{lstlisting}
  &\phantom{$\leq$}& 4 * (card (subterms b))$^\text{2}$
  $\leq$ 4 * (card (subterms (last b))
  &\phantom{$\leq$}&      + card (wits b))$^\text{2}$
  $\leq$ 4 * (card (subterms (last b))
  &\phantom{$\leq$}&      + (card (subterms (last b))$^\text{2}$)$^\text{2}$
  $\leq$ 4 * 4 * (card (subterms (last b)))$^\text{4}$.
\end{lstlisting}

@@

\begin{lstlisting}
lemma mlss_proc_branch_dom_if_wf_branch:
  assumes wf_branch b
  shows mlss_proc_branch_dom b
\end{lstlisting}

\begin{lstlisting}
lemma mlss_proc_branch_complete:
  fixes b :: "'a branch"
  assumes wf_branch b
  assumes $\neg$ mlss_proc_branch b
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ (last b)

theorem mlss_proc_complete:
  fixes $\phi$ :: 'a pset_fm
  assumes $\neg$ mlss_proc $\phi$
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ $\phi$
\end{lstlisting}

\begin{lstlisting}
lemma mlss_proc_branch_sound:
  assumes wf_branch b
  assumes $\forall$$\psi$ $\in$ set b. M $\models$ $\psi$
  shows $\neg$ mlss_proc_branch b

theorem mlss_proc_sound:
  assumes M $\models$ $\phi$
  shows $\neg$ mlss_proc $\phi$
\end{lstlisting}
